---
title: "Determining Donor and Recipient Cells"
date: "2025-05-12"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Determining Donor and Recipient Cells}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Creating a new variantCell project

If you haven't already done so, you can create a new project by using
the command:

```{r, eval=FALSE}
#Create new variantCell object
project <- variantCell$new()
```

to initialize the project.

# Determining Donor / Recipient Cells in Single Cell Data

If you need to identify which cells are donor or recipient, you can use
the process_vireo_seurat helper function (below).

```{r, eval=FALSE}
#process vireo using Seurat metadata
process_vireo_seurat = function(
  seurat_obj,
  vireo_path,
  prefix_text
)
```

Usage example:

```{r, eval=FALSE}
sample_donor_recipient <- vCell_project$process_vireo_seurat(
    sample,
    vireo_path  = "donor_ids.tsv",
    prefix_text = "prefix_text_to_add_"
)
```

This will provide output similar to below, indicating the matching cells
in the metadata, and the cell type distribution for the donors
identified by Vireo:

> Initial cell identities in Seurat object:
>
> Sample1 1152 Seurat object cells: 1152 Total Vireo cells: 1661
>
> First few Seurat rownames: Patient1_AAACCCAAGTCTAGAA-1
> Patient1_AAACGCTGTGGACAGT-1 Patient1_AAAGGATCAGTTGTTG-1
> Patient1_AAAGGATTCCTCGCAT-1 Patient1_AAAGTCCAGCACTCCG-1
> Patient1_AAAGTCCGTTGCATTG-1 First few metadata rownames:
> Patient1_AAACCCAAGTCTAGAA-1 Patient1_AAACGCTGTGGACAGT-1
> Patient1_AAAGGATCAGTTGTTG-1 Patient1_AAAGGATTCCTCGCAT-1
> Patient1_AAAGGGCTCAAAGGTA-1 Patient1_AAAGTCCAGCACTCCG-1
>
> Number of matching cells: 1152

> Cell type distribution for donor1:
>
> ```         
>   Classical Monocyte        Differentiating T            Gamma Delta T        IGSF21+ Dendritic 
>                   10                       34                       81                        9 
>      Innate Lymphoid    Intermediate Monocyte               Macrophage                     MAIT 
>                    4                        3                       29                        5 
>                 Mast                 Memory B Myeloid Dendritic Type 1 Myeloid Dendritic Type 2 
>                   13                       46                       10                       23 
>              Naive B             Naive CD4+ T             Naive CD8+ T               Neutrophil 
>                    1                       31                        2                        1 
>                   NK    Nonclassical Monocyte OLR1+ Classical Monocyte                   Plasma 
>                   14                       23                        2                       62 
> ```
>
> Cell type distribution for donor0:
>
> ```         
> Adventitial Fibroblast       Airway Smooth Muscle Alveolar Epithelial Type 1 Alveolar Epithelial Type 2 
>                      5                          1                          3                          1 
>    Alveolar Fibroblast                     Artery                      Basal                  Capillary 
>                     13                         10                         16                         38 
>     Capillary Aerocyte                   Ciliated      Differentiating Basal                     Goblet 
>                      3                          3                          1                          1 
>              Lymphatic           Microvasculature                     Mucous              Myofibroblast 
>                      1                         11                          1                          3 
>               Pericyte                     Plasma        Proliferating Basal             Proximal Basal 
>                      6                          3                          2                          3 
> Vascular Smooth Muscle                       Vein 
>                      3                         11 
> ```

In this case, the donor cells can be identified as (likely) donor0 -
with predominantly structural lung cells, while donor1, containing
mostly immune cells, is the recipient. This will be used as input to the
`addSampleData` function - to specify which donor is the
recipient/donor. This process will need to be repeated for each sample
to determine donor/recipient, based on cell type distributions.
